---
title: "Homework 4"
author: "[David Coomes](mailto:dcoomes@uw.edu)"
date: '`r format(Sys.time(), "%Y-%m-%d")`'
header-includes: 
- \usepackage{float} 
- \floatplacement{figure}{H} 
output: 
    bookdown::html_document2: 
        toc: true
        toc_float:
            collapsed: true
            smooth_scroll: false
        number_sections: false
        self_contained: true
        code_folding: hide


---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, results='hide')
```

```{r, warning=F, message=F, include=F}

library(tidyverse)
library(ggplot2)
library(purrr)

# for having DIV tags
knitr::knit_hooks$set(class = function(before, options, envir) {
  if(before){
    sprintf("<div class = '%s'>", options$class)
  }else{
    "</div>"
  }
})


```


```{r}

#creating the data
gradpain <- c(rep(0,11), rep(1,1), rep(2,2), rep(3,6), rep(4,8), rep(5,10), rep(6,8), rep(7,30), rep(8,10), rep(9,6), rep(10,2))

```



# Question 1

```{r}

gradpain %>% length


```

There are `r gradpain %>% length` graduate students in the sample

<br>

# Question 2

```{r}

gradpain %>% mean()

```

The mean of the sample is `r gradpain %>% mean`.

<br>

# Question 3


```{r}

fun_1 <- function(d.vec, n) {
  s <- sample(x=d.vec, size=n, replace=TRUE)
  list(vec_length=length(s),
       vec_mean=mean(s))
}

```


```{r}

fun_1(gradpain, length(gradpain))
x <- fun_1(gradpain, length(gradpain))


```

```{r}

set.seed(7)
out <- NULL
for (i in 1:100){
  mean <- fun_1(d.vec=gradpain, n=length(gradpain))$vec_mean
  out <- c(out, mean)
}

mean(out)
sd(out)


```


The mean of the 100 means `r round(mean(out), 2)`.

<br>

# Question 4

The standard deviation of the 100 means is `r round(sd(out), 2)`. 

<br>

# Question 5


```{r}

set.seed(7)
fun_2 <- function(d.vec, n, num.samples){
  out <- NULL
  for (i in 1:num.samples){
    s <- sample(x=d.vec, size=n, replace=TRUE)
    s_size <- length(s)
    s_mean <- mean(s)
    out <- c(out, s_mean)
  }
  ci_hi <- mean(out) + qnorm(0.975)*(sd(out)/sqrt(s_size))
  ci_low <- mean(out) - qnorm(0.975)*(sd(out)/sqrt(s_size))
  ci <- list(ci_low, ci_hi)
  mylist <- list(s_size, num.samples, out, mean(out), sd(out), ci)
  #mylist <- list(s_size, num.samples, out, mean(out), sd(out), confint(out))
  return(mylist)
}

myfun <- fun_2(gradpain, length(gradpain), 100)


```




The mean of means is `r round(myfun[[4]], 2)`.

<br>

# Question 6

The standard deviation of means is `r round(myfun[[5]], 2)`.

<br>

# Question 7

The 95% confidence interval around the means is `r round(myfun[[6]][[1]], 2)` - `r round(myfun[[6]][[2]], 2)`.

<br>

# Code

```{r comment='', results='show'}
cat(readLines("dcoomes_hw_04.Rmd"), sep = '\n')
```


