---
title: "Homework 2"
author: "David Coomes"
date: "1/19/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



Starting out by loading libraries.

```{r, warning=FALSE, message=FALSE}

library(tidyverse)
library(kableExtra)
library(janitor)

```



### Question 5.1

Bringing in data. 

```{r}

data <- iris

```


Creating one table for each species that includes all variables from the Iris data set. 

```{r}

data %>%
  filter(Species=="setosa") %>%
  kable(align="c") %>% kable_styling(bootstrap_options = c("striped", "condensed"), full_width=F,  position="left")

data %>% 
  filter(Species=="versicolor")  %>%
  kable(align="c") %>% kable_styling(bootstrap_options = c("striped", "condensed"), full_width=F,  position="left")

data %>% 
  filter(Species=="virginica")  %>%
  kable(align="c") %>% kable_styling(bootstrap_options = c("striped", "condensed"), full_width=F,  position="left")

```

<br>

### Question 5.2 

Setting up a table including only spp. *virginica* that lists whether the sepal length and width of that individual is larger than the mean of the sample. 

```{r}

data %>% 
  filter(Species=="virginica") %>%
  mutate(width_mean=mean(Sepal.Width),
         length_mean=mean(Sepal.Length),
         wider_than_mean = ifelse(Sepal.Width>width_mean, "Yes", "No"),
         longer_than_mean = ifelse(Sepal.Length>length_mean, "Yes", "No")) %>%
  select(Sepal.Length, Sepal.Width, wider_than_mean, longer_than_mean) %>%
  kable(align="c") %>% kable_styling(bootstrap_options = c("striped", "condensed"), full_width=F, position = "left") 

```


<br>

### Question 5.3 

Reading in health data set. 

```{r, warning=FALSE, results='hide'}

data <- readstata13::read.dta13("data/AHwave1_v1.dta")
DT::datatable(data.frame(colname = names(data), label = attributes(data)$var.labels))
data$h1gh1 %>% unique()

```

Creating variable that classifies health as: Best (excellent, very good) or Worst (good, fair, poor). 

```{r}

data <- data %>%
  mutate(health = ifelse((h1gh1=="(1) Excellent" | h1gh1 == "(2) Very good"), "Best", 
         ifelse((h1gh1 == "(3) Good" | h1gh1 == "(4) Fair" | h1gh1 == "(5) Poor"), "Worst", NA))) 

#checking to make sure it is created correctly
#table(data$h1gh1, data$health, useNA = "ifany")

```


<br>

### Question 5.4

Table of the the count and proportion of sample who had Best (excellent or very good) and Worst (good, fair, or poor) health.

```{r}

tabyl(data$health, sort=TRUE) %>%
  kable(align = "c") %>% kable_styling(bootstrap_options = c("striped", "condensed"), full_width=F, position = "left") 

```


<br>

### Question 5.5

Table of the the count and proportion of sample who had Best (excellent or very good) and Worst (good, fair, or poor) health by race for white and black respondents.


```{r}

data %>%
  filter(h1gi6a=="(1) Marked") %>%
  mutate(health_white=health) %>%
  tabyl(health_white) %>%
  kable(align = "c") %>% kable_styling(bootstrap_options = c("striped", "condensed"), full_width=F, position = "left")

data %>%
  filter(h1gi6b=="(1) Marked") %>%
  mutate(health_black=health) %>%
  tabyl(health_black) %>%
  kable(align = "c") %>% kable_styling(bootstrap_options = c("striped", "condensed"), full_width=F, position = "left")


```








